The lattice of acyclic pipe dreams

### 2023-01-11 acyclicFacets/draft/latticePipeDreams.tex
a basic introduction

% theorems
\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{conjecture}[theorem]{Conjecture}

@@

\begin{abstract}
We show that for any permutation~$\omega$, the increasing flip graph on acyclic pipe dreams with permutation~$\omega$ is a lattice quotient of the interval~$[e,\omega]$ of the weak order.
We then discuss conjectural generalizations of this result to acyclic facets of subword complexes on arbitrary finite Coxeter groups.
\end{abstract}

\maketitle

@@

The purpose of this note is to show that for any permutation~$\omega$, the increasing flip graph on acyclic pipe dreams with permutation~$\omega$ is a lattice quotient of the interval~$[e,\omega]$ of the weak order.
It motivates \cref{conj:Coxeter} which is the long-term perspective of this work.
The approach and proofs developed here are direct adaptations from~\cite[Sect.~1]{Pilaud-BrickAlgebra}.
\vincent{Include an introduction}
\vincent{cite \cite{BergeronBilley,KnutsonMiller-subwordComplex}}

The \defn{weak order} on the permutations of~$[n]$ is defined as the inclusion order of inversion sets, where an inversion of a permutation~$\pi$ is a pair of values~$i, j \in [n]$ such that~$i < j$ while~${\pi^{-1}(i) > \pi^{-1}(j)}$.
The cover relations in the weak order correspond to switching pairs of consecutive values in permutations.
The weak order is known to be a lattice \cite{GuilbaudRosenstiehl,Bjorner}.
For two permutations~$\sigma < \tau$, we denote by~$[\sigma, \tau]$ the weak order interval between~$\sigma$ and~$\tau$, which is again a lattice.

@@

\subsection{Pipe dream quotient}
\label{subsec:}

@@

\newpage
\section{Subword complexes}

The objective of this section is to partially extend the results about the lattice structure of acyclic pipe dreams, presented in \cref{sec:latticeAcyclicPipeDreams}, to the wider context of subword complexes of finite type.
Before doing so, we recall some preliminaries about finite Coxeter groups and subword complexes.
We refer to~\cite{BjornerBrenti, Humphreys} for more details on Coxeter groups.

@@

Two facets $I,J\in\subwordComplex(Q,w)$ are said to be connected by a \defn{flip} if they differ one element:
\[J=I\smallsetminus \{i\} \cup\{j\}\]
for some $i\in I$ and $j\in J$ with $i\neq j$.
If $i<j$, the flip from $I$ to $J$ is called an \defn{increasing flip}.

>>>>

% theorems
\newtheorem*{theorem*}{Theorem}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{conjecture}[theorem]{Conjecture}

@@

\begin{abstract}
We show that for any permutation~$\omega$, the increasing flip graph on acyclic pipe dreams with exiting permutation~$\omega$ is a lattice quotient of the interval~$[e,\omega]$ of the weak order.
We then discuss conjectural generalizations of this result to acyclic facets of subword complexes on arbitrary finite Coxeter groups.
\end{abstract}

\vspace*{.2cm}
\maketitle

@@

The weak order is the lattice on permutations of~$[n]$ whose cover relations correspond to switching pairs of consecutive values in permutations.
The Tamari lattice is the lattice on binary trees with~$n$ internal nodes whose cover relations correspond to right rotations in binary trees.
The Tamari lattice is known to be the lattice quotient of the weak order by the sylvester congruence, defined as the equivalence relation on permutations of~$[n]$ whose equivalence classes are the sets of linear extensions of binary trees (labeled in inorder and oriented towards their leaves).

This paper develops a similar framework for acyclic pipe dreams.
Pipe dreams were introduced by N.~Bergeron and S.~Billey in~\cite{BergeronBilley} to compute Schubert polynomials and later revisited in the context of Gr\"obner geometry by A.~Knutson and E.~Miller~\cite{KnutsonMiller-GroebnerGeometry}, who coined the name \emph{pipe dreams}.
They have important connections and applications to various areas related to Schubert calculus and Schubert varieties~\cite{LascouxSchutzenberger-PolynomesSchubert, LascouxSchutzenberger-SchubertLittlewoudRichardson}. 
A pipe dream is an arrangement of pipes in the triangular shape, each entering along the vertical side and exiting along the horizontal side.
They are grouped according to their exiting permutation, given by the order in which the pipes appear along the horizontal axis.
The pipes in a pipe dream have some crossings and some contacts, and the contact graph records all contacts of a pipe dream.
A flip in a pipe dream exchanges a contact between with a crossing between two pipes, and the flip is increasing when the contact is southeast of the crossing involved in the flip.
The acyclic pipe dreams are the pipe dreams with an acyclic contact graph, and they naturally appear in the study of brick polytopes~\cite{PilaudSantos-brickPolytope}.
A brief recollection on pipe dreams is given in \cref{sec:preliminaries}.

In the core \cref{sec:latticeAcyclicPipeDreams} of this paper, we show that for any permutation~$\omega$,
\begin{itemize}
\item the sets of linear extensions of the acyclic pipe dreams with exiting permutation~$\omega$ form a partition of the interval~$[e,\omega]$ of the weak order (\cref{subsec:linearExtensions}),
\item the equivalence relation defined by this partition is a lattice congruence of~$[e, \omega]$, that we call the pipe dream congruence (\cref{subsec:pipeDreamCongruence}),
\item the Hasse diagram of the corresponding lattice quotient is the increasing flip graph on acyclic pipe dreams with exiting permutation~$\omega$ (\cref{subsec:pipeDreamQuotient}).
\end{itemize}
In summary, we obtain the following statement.

\begin{theorem*}
For any permutation~$\omega$, the increasing flip graph on acyclic pipe dreams with exiting permutation~$\omega$ is the Hasse diagram of the lattice quotient of the interval~$[e,\omega]$ of the weak order by the pipe dream congruence of~$\omega$.
\end{theorem*}

Note that we recover the connection between the weak order and the Tamari lattice by the sylvester congruence for a well-chosen exiting permutation~$\omega$.

We then explore in \cref{sec:furtherTopics} some natural further topics.
We describe two algorithms to compute the unique acyclic pipe dream whose linear extensions contain a given permutation generalizing the binary tree insertion map on permutations (\cref{subsec:sweepingAlgorithm,subsec:insertionAlgorithm}), we describe the pipe dream congruence as the transitive closure of a local rewriting rule generalizing that of the sylvester congruence (\cref{subsec:rewritingRule}), and we describe a natural commutative diagram of lattice morphism generalizing the connection between the recoil map and the binary tree insertion map on permutations and the canopy on binary trees (\cref{subsec:canopy}).

Finally, we discuss in \cref{sec:subwordComplexes} a conjectural extension of our results to subword complexes in finite Coxeter groups~\cite{KnutsonMiller-subwordComplex}.

@@

\subsection{Pipe dream quotient}
\label{subsec:pipeDreamQuotient}

@@

\newpage
\section{Subword complexes}
\label{sec:subwordComplexes}

The objective of this section is to partially extend the results about the lattice structure of acyclic pipe dreams, presented in \cref{sec:latticeAcyclicPipeDreams}, to the wider context of subword complexes in finite Coxeter groups.
Before doing so, we recall some preliminaries about finite Coxeter groups and subword complexes.
We refer to~\cite{BjornerBrenti, Humphreys} for more details on Coxeter groups.

@@

Two facets $I,J \in \subwordFacets(Q,w)$ are said to be connected by a \defn{flip} if they differ one element:
\[J=I\smallsetminus \{i\} \cup\{j\}\]
for some $i\in I$ and $j\in J$ with $i\neq j$.
If $i<j$, the flip from $I$ to $J$ is called an \defn{increasing flip}.
